---
title: "pager_encode"
author: "Attri Ghosh"
date: "2024-02-21"
output: html_document
---
## This R markdown is used to encode sample genotype datasets using PAGER (Phenotype-Adjusted Genotype Encoding and Ranking). 
Dataset 1 - Continuous Phenotype
Dataset 2 - Discrete Phenotype 

## STEP 1 - Import all the required packages

```{r setup, include=FALSE}
# install all the required packages
library(dplyr)
library(caret)
```


## STEP 2 - Define PAGER ENCODING function

```{r setup, include=TRUE}
# Function to perform PAGER encoding
pager_encode <- function(genotypes, phenotype) {
  
  # Create an empty dataframe to store the PAGER encoded genotypes
  pager_encoded_genotypes <- data.frame(matrix(ncol = ncol(genotypes), nrow = nrow(genotypes)))
  colnames(pager_encoded_genotypes) <- colnames(genotypes)
  
  # Create an empty dataframe to store the SNPs with missing genotypes
  missing_genotypes_df <- data.frame(SNP_NAME = character(), MISSING_GENOTYPES = character(), stringsAsFactors = FALSE)
  
  # Iterate through the columns of the genotypes dataframe to calculate the PAGER encoding for each SNP
  pager_encoding_info <- list() # Store encoding information
  
  for (column in colnames(genotypes)) {
    
    # Extract the single SNP from the genotypes dataframe
    snp <- genotypes[[column]]
    
    # Create a dataframe with the single SNP and the phenotype
    snp_df <- data.frame(genotype = as.numeric(snp), phenotype = as.numeric(phenotype))
    
    # Identify unique genotypic classes, excluding NAs
    unique_genotypes <- na.omit(unique(snp_df$genotype))
    
    # Check for missing genotypic classes
    missing_genotypes <- setdiff(c(0, 1, 2), unique_genotypes)
    
    # Log missing genotypes
    if (length(missing_genotypes) > 0) {
      cat(paste("Missing genotypes for", column, ":", toString(missing_genotypes), "\n"))
      snp_missing_genotype_df <- data.frame(SNP_NAME = column, MISSING_GENOTYPES = toString(missing_genotypes), stringsAsFactors = FALSE)
      missing_genotypes_df <- bind_rows(missing_genotypes_df, snp_missing_genotype_df)
    }
    
    # Check if only one genotypic class is present
    if (length(missing_genotypes) == 2) {
      pager_encoded_genotypes[[column]] <- snp
      next
    }
    
    # Calculate phenotypic mean per genotypic class for non-NA genotypes
    geno_aggregations <- snp_df %>%
      filter(!is.na(genotype)) %>%
      group_by(genotype) %>%
      summarise(mean_phenotype = mean(phenotype)) %>%
      ungroup() %>%
      mutate(genotype = as.numeric(genotype))
    
    # Use the PAGER formula
    anchor_mean <- geno_aggregations[which.min(geno_aggregations$mean_phenotype), 'mean_phenotype'] # anchor = 0 when all three genotypes are present
    geno_aggregations$rel_dist <- as.numeric(geno_aggregations$mean_phenotype) - as.numeric(anchor_mean)
    geno_aggregations$normalized_rel_dist <- (geno_aggregations$rel_dist - min(geno_aggregations$rel_dist)) / (max(geno_aggregations$rel_dist) - min(geno_aggregations$rel_dist))
    
    # Assign the normalized values to the corresponding SNP column
    pager_encoded_genotypes[[column]] <- geno_aggregations$normalized_rel_dist[match(snp, geno_aggregations$genotype)]
    
    # Store encoding information for application to test set
    pager_encoding_info[[column]] <- geno_aggregations
  }
  
  # Return the PAGER encoded genotypes, missing genotypes dataframe, and encoding information
  return(list(pager_encoded_genotypes, missing_genotypes_df, pager_encoding_info))
}
```

## STEP 3 - Function to apply PAGER encoding to a new dataset
```{r setup, include=TRUE}
# Function to apply PAGER encoding to a new dataset
apply_pager_encoding <- function(genotypes, encoding_info) {
  pager_encoded_genotypes <- data.frame(matrix(ncol = ncol(genotypes), nrow = nrow(genotypes)))
  colnames(pager_encoded_genotypes) <- colnames(genotypes)
  
  for (column in colnames(genotypes)) {
    snp <- genotypes[[column]]
    encoding <- encoding_info[[column]]
    if (!is.null(encoding)) {
      pager_encoded_genotypes[[column]] <- encoding$normalized_rel_dist[match(snp, encoding$genotype)]
    } else {
      pager_encoded_genotypes[[column]] <- snp
    }
  }
  return(pager_encoded_genotypes)
}
```

## STEP 4 - PAGER encode Dataset 1 - Continuous Phenotype
```{r setup, include=TRUE}
# Read the data from the csv file
data_continuous_phenotype <- read.csv('./PAGER/data/sample_data_cont.csv')

# Split into genotypes and phenotype
genotype_columns <- data_continuous_phenotype[, -ncol(data_continuous_phenotype)]
phenotype <- data_continuous_phenotype[, ncol(data_continuous_phenotype)]

# Split the data into training and testing sets
set.seed(123) # for reproducibility
train_index <- createDataPartition(phenotype, p = 0.7, list = FALSE)
train_genotypes <- genotype_columns[train_index, ]
train_phenotype <- phenotype[train_index]
test_genotypes <- genotype_columns[-train_index, ]
test_phenotype <- phenotype[-train_index]

# Encode the training dataset
train_result <- pager_encode(train_genotypes, train_phenotype)

# Apply the PAGER encoding to the test dataset
test_encoded_genotypes <- apply_pager_encoding(test_genotypes, train_result[[3]])

# Combine the PAGER encoded genotypes with the phenotypes
train_pager_encoded <- cbind(train_result[[1]], train_phenotype)
test_pager_encoded <- cbind(test_encoded_genotypes, test_phenotype)

# Write the encoded datasets to CSV
write.csv(train_pager_encoded, file = 'path/to/folder/train_pager_cont_encoded.csv', row.names = FALSE)
write.csv(test_pager_encoded, file = 'path/to/folder/test_pager_cont_encoded.csv', row.names = FALSE)
write.csv(train_result[[2]], file = 'path/to/folder/missing_cont_genotypes.csv', row.names = FALSE)
```

## STEP 5 - PAGER encode Dataset 2 - Discrete Phenotype
```{r setup, include=TRUE}
# Read the data from the csv file
data_discrete_phenotype <- read.csv('./PAGER/data/sample_data_dis.csv')

# Split into genotypes and phenotype
genotype_columns <- data_discrete_phenotype[, -ncol(data_discrete_phenotype)]
phenotype <- data_discrete_phenotype[, ncol(data_discrete_phenotype)]

# Split the data into training and testing sets
set.seed(123) # for reproducibility
train_index <- createDataPartition(phenotype, p = 0.7, list = FALSE)
train_genotypes <- genotype_columns[train_index, ]
train_phenotype <- phenotype[train_index]
test_genotypes <- genotype_columns[-train_index, ]
test_phenotype <- phenotype[-train_index]

# Encode the training dataset
train_result <- pager_encode(train_genotypes, train_phenotype)

# Apply the PAGER encoding to the test dataset
test_encoded_genotypes <- apply_pager_encoding(test_genotypes, train_result[[3]])

# Combine the PAGER encoded genotypes with the phenotypes
train_pager_encoded <- cbind(train_result[[1]], train_phenotype)
test_pager_encoded <- cbind(test_encoded_genotypes, test_phenotype)

# Write the encoded datasets to CSV
write.csv(train_pager_encoded, file = 'path/to/folder/train_pager_dis_encoded.csv', row.names = FALSE)
write.csv(test_pager_encoded, file = 'path/to/folder/test_pager_dis_encoded.csv', row.names = FALSE)
write.csv(train_result[[2]], file = 'path/to/folder/missing_dis_genotypes.csv', row.names = FALSE)
```

